---
title: "Data Quality Control"
author: "Denisse Fierro Arcos"
date: "2023-09-23"
output:
  github_document:
    toc: true
    html_preview: false
---

# Quality Control (QC) workflow

In this document, we will apply a quality control workflow to the [Antarctic Phipoda](https://ipt.biodiversity.aq/resource?r=2438_mista_ant_phipoda#anchor-downloads) dataset available from the IPT.

## Loading libraries

```{r message = F, warning = F}
#Data manipulation and plotting
library(tidyverse)
#Downloading data
library(curl)
```
  
## Downloading data from IPT
  
First, we will provide the path to the folder where where we will store our data. We will call this folder `Data`, and it will be located in the main repository folder. If this folder does not yet exist, we will create it. 
  
```{r}
#Defining location of data folder
folder_out <- "../Data"

#Checking if folder exists, if not, one will be created
if(dir.exists(folder_out) == F){
  dir.create(folder_out)
}
```
  
We are now ready to download our data from the IPT using the `curl` package.  
  
```{r}
#Create full path to save downloaded data
path_out <- file.path(folder_out, "Antarctic_Phipoda.zip")

#Downloading files from IPT
curl_download("https://ipt.biodiversity.aq/archive.do?r=2438_mista_ant_phipoda&v=1.1", 
              destfile = path_out)

#Unzipping files into the data folder
unzip(path_out, exdir = folder_out)

#Finally, we will get a list of all files downloaded
file_list <- list.files(path = folder_out, full.names = T)
#We can check results now
file_list
```
  
We can see that this dataset has a relatively simple structure as it only contains one core file: `occurrence.txt`, which contains information about where Antarctic Phipoda has been recorded. In this case, we can go ahead and load the `occurrence.txt` into our environment. The `xml` files provide metadata, which will not be looking into in this repository.  
  
## Loading data 
  
```{r}
ant_phipoda <- read_delim(str_subset(file_list, ".*txt"), delim = "\t")
#Checking contents
glimpse(ant_phipoda)
```
```{r}
library(rgif)
y <- occ_download(pred("datasetKey", "667e0044-3ff1-4365-8ecf-e637c7b31fa4"))
occ_download_wait('0005146-230918134249559')
d <- occ_download_get('0005146-230918134249559') %>%
    occ_download_import()

```


