---
title: "02_data_plots"
author: "Denisse Fierro Arcos"
date: "2023-09-24"
output:
  github_document:
    toc: true
    html_preview: false
---

# Plotting Antarctic Phipoda data

The first notebook in this repository shows the quality control workflow applied to the Antarctic Phipoda dataset available in GBIF. Here, we will use the cleaned dataset to create some plots.  
  
## Loading libraries
  
```{r message=F, warning=FALSE}
library(tidyverse)
library(treemap)
library(ggOceanMaps)
library(ggspatial)
```
  
## Loading data
Here we will load the cleaned dataset produced in the previous notebook.  
  
```{r}
ant_phi <- read_csv("../Data/Antarctic_Phipoda_GBIF_QC.csv")
#Checking structure
glimpse(ant_phi)
```
  
## Plotting data
The first plot will show the number of individuals per family of Antarctic Phipoda included in the dataset.  
  
```{r}
ant_phi %>% 
  select(family, verbatimScientificName) %>% 
  count(family, verbatimScientificName) %>% 
  ungroup() %>% 
  complete(verbatimScientificName, nesting(family)) %>% 
  group_by(family) %>%
  mutate(fam = sum(n, na.rm = T)) %>% 
  unite("family", family, fam, sep = "\n") %>% 
  treemap(., index = c("family"), vSize = "n",
          fontsize.labels = 10, fontcolor.labels = "black",
          align.labels = c("center", "center"),
          overlap.labels = 0.5, border.col = "grey", 
          palette = "Dark2", 
          title = "Number of individual Antarctic Phipoda sampled per family in the Southern Ocean")
```
    
Now we can create a map of observations. We will create a map with a South Pole Stereographic projection using the `ggOceanMaps` and `ggspatial` libraries.
  
```{r}
#First, we state the northern most limit for our map
#We also overlay the bathymetry
basemap(limits = -30, bathymetry = T)+
  #Now we overlay our data points
  geom_spatial_point(data = ant_phi, color = "red", crs = 4326,
                     aes(x = decimallongitude, y = decimallatitude))

```
  
Any figures can be saved into memory using the `ggsave` function from `ggplot2`.  





